Thesis Draft

### 2017-10-15 thesis/draft.Rmd
fixed test data MARE calculation to usse false-zero count corrected data

---
title: "Thesis Draft"
author: "Gavin Chin"
date: "21 September 2017"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

@@

```{r, fig.width=4, fig.height=4, fig.align='center', fig.cap = "Map showing the algorithm selecting Flinders St Station Underpass and Sandridge Bridge as neighbours to be used to predict the pedestrian counts at Southbank"}
ped_loc <- read_csv("../data/Pedestrian_sensor_locations.csv")
lats <- ped_loc$Latitude
lons <- ped_loc$Longitude
ids <- ped_loc$`Sensor Description`

@@

sens_lines <- data.frame(lon = sens_lines_lon,
                         lat = sens_lines_lat)
options(na.action = 'na.omit')
ggmap(melb3) + geom_line(aes(x = lon, y = lat),
                        data = sens_lines, size = 2.5,
                        colour = viridis(3)[2], alpha = 0.5) +
               geom_point(aes(x = lons, y = lats, colour = cl),
                          size = 8,
                          data = loc) +
               scale_colour_viridis(discrete = TRUE,
                                    name = "Sensor NA Prop.",
                                    begin = 0.5, end = 0) +
               geom_point(aes(x = loc["Southbank",]$lons,
                              y = loc["Southbank",]$lats),
                          size = 16, colour = viridis(3)[2]) +
               geom_label(aes(label = ids, x = lons, y = lats,
                              fill = cl),
                         fontface = 'bold', colour = 'white',
                         data = loc, size = 3.5,
                         nudge_y = -0.000325) +
              scale_fill_viridis(discrete = TRUE,
                                  name = "Sensor NA Prop.",
                                  begin = 0.5, end = 0) +
              xlab(NULL) + ylab(NULL) +
              theme(plot.title = element_text(hjust = 0.5),
                    plot.caption = element_text(hjust = 0.5),
                    axis.text.x = element_blank(),
                    axis.text.y = element_blank(),
                    axis.ticks = element_blank(),
                    legend.position = c(.8,0.2))
```

@@

MARE_base_ts <- numeric()
for (i in unique(ped_data$Sensor)){
  MARE_base_ts[i] <- mean(abs(test_base[[i]] - unlist(ori_ts[, i])), na.rm = T) / mean(unlist(ori_ts[, i]), na.rm = T)
}

MARE_ts <- numeric()
for (i in unique(ped_data$Sensor)){
  MARE_ts[i] <- mean(abs(test_fit[[i]] - unlist(ori_ts[, i])), na.rm = T) / mean(unlist(ori_ts[, i]), na.rm = T)
}
```

Again, we start by comparing the fit at a sensor location with a small proportion of missing values, Vic Arts Centre:
```{r, fig.height=4, fig.width=9, fig.cap = "A plot of the out-of-sample (2017) Actual vs Predicted pedestrian counts at Melbourne Central (sensor location with small proportion of missing values) for the simple GLM model and the improved GLM model. The dashed line is $x=y$, representing a reference for perfect fit, while solid line is the linear fit of the points."}
p1 <- ggplot(data = NULL, aes(x = ori_ts[, "Melbourne Central"],
           y = test_base$`Melbourne Central`)) +
           geom_point(alpha = 0.05) +
           geom_abline(aes(slope = 1, intercept = 0),
                       colour = viridis(3)[1],
                       alpha = 0.7, size = 1.25,
                       linetype = 2) +
           geom_smooth(method = "lm", size = 1.25, alpha = 0.7, 
                       colour = viridis(3)[3]) +
           theme_minimal() + 
           theme(aspect.ratio = 1,
                 plot.title = element_text(hjust = 0.5)) +
           scale_x_continuous(limits = c(0, 4000)) +
           scale_y_continuous(limits = c(0, 4000)) +
           xlab("Actual Pedestrian Count") +
           ylab("Predicted Pedestrian Count") +
           ggtitle("Simple GLM Model: Actual v. Predicted \n at Melbourne Central")
p2 <- ggplot(data = NULL, aes(x =ori_ts[, "Melbourne Central"],
           y = test_fit$`Melbourne Central`)) +
           geom_point(alpha = 0.05) +
           geom_abline(aes(slope = 1, intercept = 0),
                       colour = viridis(3)[1],
                       alpha = 0.7, size = 1.25,
                       linetype = 2) +
           geom_smooth(method = "lm", size = 1.25, alpha = 0.7, 
                       colour = viridis(3)[3]) +
           theme_minimal() + 
           theme(aspect.ratio = 1,
                 plot.title = element_text(hjust = 0.5)) +
           scale_x_continuous(limits = c(0, 4000)) +
           scale_y_continuous(limits = c(0, 4000)) +
           xlab("Actual Pedestrian Count") +
           ylab("Predicted Pedestrian Count") +
           ggtitle("Improved GLM Model: Actual v. Predicted \n at Melbourne Central")
grid.arrange(p1, p2, nrow = 1)
```

@@

```{r}
MARE_base_ts2 <- MARE_base_ts[setdiff(unique(ped_data$Sensor),
                                      c("Spencer St-Collins St (South)",
                                        "Spencer St-Collins St (North)"))]
MARE_ts2 <- MARE_ts[setdiff(unique(ped_data$Sensor),
                                      c("Spencer St-Collins St (South)",
                                        "Spencer St-Collins St (North)"))]

@@

```{r, fig.height=3.5, fig.width=9, fig.cap = paste("Distribution of out-of-sample (2017 data) MARE by imputation model, excluding Spencer St-Collins St (North)/(South) sensors")}

@@

Spencer St-Collins St (North)/(South) was omitted from the the plots above as the out-of-sample MARE is extremely high at `r round(MARE_ts["Spencer St-Collins St (North)"]*100, 2)`% and `r round(MARE_ts["Spencer St-Collins St (South)"]*100, 2)`% respectively.

As these sensor locations were both classified as locations with a large proportion of missing values, it would appear that the relationship between these locations and their neighbouring sensors has changed. This is because the in-smaple MARE at these sensor locations were quite acceptable at `r round(MARE_tr["Spencer St-Collins St (North)"]*100, 2)`% and `r round(MARE_tr["Spencer St-Collins St (South)"]*100, 2)`% respectively.

@@

Another way we can perform cross-validation is to remove data at a sensor, and evaulate the predictions made. This will allow us to evaluate the imputation model's performance within sample, which is the primary concern with the imputation process.  
We select a sensor which has a small proportion of missing values, Bourke Street Mall (South), which has `r round((1 - na_prop["Bourke Street Mall (South)"])*100, 3)`% of data available. As observed in the data, missing data is present in different ways: random, short periods of missing data (no more that 6 hours) or long periods of missing data (multiple months). We will simulate both types of missingness at this sensor location.

@@

```{r, fig.width = 8, fig.height=2, fig.cap = "Plot of simulated missingness method 1 at Bourke Street Mall (South)"}
ggplot() + geom_line(aes(x = dfa2$Date_Time, y = mall_missing_1)) +
           labs(y = "Pedestrian Count", x = "Time and Date",
                title = "Simulated Missingness 1") +
          theme_minimal() + theme(plot.title = element_text(hjust = 0.5))
```

@@

```{r, fig.width = 8, fig.height=2, fig.cap = "Plot of simulated missingness method 2 at Bourke Street Mall (South)"}
ggplot() + geom_line(aes(x = dfa2$Date_Time, y = mall_missing_2)) +
           labs(y = "Pedestrian Count", x = "Time and Date",
                title = "Simulated Missingness 2") +
          theme_minimal() + theme(plot.title = element_text(hjust = 0.5))
```

>>>>

---
title: "Thesis Draft"
author: "Gavin Chin"
date: "21 September 2017"
output:
  pdf_document: default
  html_document: default
  word_document: default
fontsize: 12pt
header-includes:
- \usepackage{setspace}
- \doublespacing
---

@@

```{r, fig.width=3, fig.height=3, fig.align='center', fig.cap = "Map showing the algorithm selecting Flinders St Station Underpass and Sandridge Bridge as neighbours to be used to predict the pedestrian counts at Southbank"}
ped_loc <- read_csv("../data/Pedestrian_sensor_locations.csv")
lats <- ped_loc$Latitude
lons <- ped_loc$Longitude
ids <- ped_loc$`Sensor Description`

@@

sens_lines <- data.frame(lon = sens_lines_lon,
                         lat = sens_lines_lat)
options(na.action = 'na.omit')
ggmap(melb3) + geom_line(aes(x = lon, y = lat),
                        data = sens_lines, size = 2,
                        colour = viridis(3)[2], alpha = 0.5) +
               geom_point(aes(x = lons, y = lats, colour = cl),
                          size = 6,
                          data = loc) +
               scale_colour_viridis(discrete = TRUE,
                                    name = "Sensor NA Prop.",
                                    begin = 0.5, end = 0) +
               geom_point(aes(x = loc["Southbank",]$lons,
                              y = loc["Southbank",]$lats),
                          size = 12, colour = viridis(3)[2]) +
               geom_label(aes(label = ids, x = lons, y = lats,
                              fill = cl),
                         fontface = 'bold', colour = 'white',
                         data = loc, size = 3,
                         nudge_y = -0.000325) +
              scale_fill_viridis(discrete = TRUE,
                                  name = "Sensor NA Prop.",
                                  begin = 0.5, end = 0) +
              xlab(NULL) + ylab(NULL) +
              theme(plot.title = element_text(hjust = 0.5),
                    plot.caption = element_text(hjust = 0.5),
                    axis.text.x = element_blank(),
                    axis.text.y = element_blank(),
                    axis.ticks = element_blank(),
                    legend.position = c(.8,0.2))
```

@@

MARE_base_ts <- numeric()
for (i in unique(ped_data$Sensor)){
  MARE_base_ts[i] <- mean(abs(test_base[[i]] - unlist(ori_ts2[, i])), na.rm = T) / mean(unlist(ori_ts2[, i]), na.rm = T)
}

MARE_ts <- numeric()
for (i in unique(ped_data$Sensor)){
  MARE_ts[i] <- mean(abs(test_fit[[i]] - unlist(ori_ts2[, i])), na.rm = T) / mean(unlist(ori_ts2[, i]), na.rm = T)
}
```

Again, we start by comparing the fit at a sensor location with a small proportion of missing values, Melbourne Central:

```{r, fig.height=4, fig.width=9, fig.cap = "A plot of the out-of-sample (2017) Actual vs Predicted pedestrian counts at Melbourne Central (sensor location with small proportion of missing values) for the simple GLM model and the improved GLM model. The dashed line is $x=y$, representing a reference for perfect fit, while solid line is the linear fit of the points."}
p1 <- ggplot(data = NULL, aes(x = ori_ts[, "Melbourne Central"],
           y = test_base$`Melbourne Central`)) +
           geom_point(alpha = 0.05) +
           geom_abline(aes(slope = 1, intercept = 0),
                       colour = viridis(3)[1],
                       alpha = 0.7, size = 1.25,
                       linetype = 2) +
           geom_smooth(method = "lm", size = 1.25, alpha = 0.7, 
                       colour = viridis(3)[3]) +
           theme_minimal() + 
           theme(aspect.ratio = 1,
                 plot.title = element_text(hjust = 0.5)) +
           scale_x_continuous(limits = c(0, 4000)) +
           scale_y_continuous(limits = c(0, 4000)) +
           xlab("Actual Pedestrian Count") +
           ylab("Predicted Pedestrian Count") +
           ggtitle("Simple GLM Model: Actual v. Predicted \n at Melbourne Central")
p2 <- ggplot(data = NULL, aes(x =ori_ts[, "Melbourne Central"],
           y = test_fit$`Melbourne Central`)) +
           geom_point(alpha = 0.05) +
           geom_abline(aes(slope = 1, intercept = 0),
                       colour = viridis(3)[1],
                       alpha = 0.7, size = 1.25,
                       linetype = 2) +
           geom_smooth(method = "lm", size = 1.25, alpha = 0.7, 
                       colour = viridis(3)[3]) +
           theme_minimal() + 
           theme(aspect.ratio = 1,
                 plot.title = element_text(hjust = 0.5)) +
           scale_x_continuous(limits = c(0, 4000)) +
           scale_y_continuous(limits = c(0, 4000)) +
           xlab("Actual Pedestrian Count") +
           ylab("Predicted Pedestrian Count") +
           ggtitle("Improved GLM Model: Actual v. Predicted \n at Melbourne Central")
grid.arrange(p1, p2, nrow = 1)
```

@@

```{r}
MARE_base_ts2 <- MARE_base_ts[setdiff(unique(ped_data$Sensor),
                                      c("Spencer St-Collins St (South)",
                                        "Spencer St-Collins St (North)",
                                        "Flinders St-Swanston St (West)"))]
MARE_ts2 <- MARE_ts[setdiff(unique(ped_data$Sensor),
                                      c("Spencer St-Collins St (South)",
                                        "Spencer St-Collins St (North)",
                                        "Flinders St-Swanston St (West)"))]

@@

```{r, fig.height=3.5, fig.width=9, fig.cap = paste("Distribution of out-of-sample (2017 data) MARE by imputation model, excluding Spencer St-Collins St (North)/(South) and Flinders St-Swanston St (West) sensors")}

@@

Spencer St-Collins St (North)/(South) and Flinders St-Swanston St (West) was omitted from the the plots above as the out-of-sample MARE is extremely high at `r round(MARE_ts["Spencer St-Collins St (North)"]*100, 2)`%, `r round(MARE_ts["Spencer St-Collins St (South)"]*100, 2)`% and `r round(MARE_ts["Flinders St-Swanston St (West)"]*100, 2)`% respectively.

As these sensor locations were both classified as locations with a large proportion of missing values, it would appear that the relationship between these locations and their neighbouring sensors has changed. This is because the in-saple MARE at these sensor locations were quite acceptable at `r round(MARE_tr["Spencer St-Collins St (North)"]*100, 2)`%, `r round(MARE_tr["Spencer St-Collins St (South)"]*100, 2)`% and `r round(MARE_tr["Flinders St-Swanston St (West)"]*100, 2)`% respectively.

@@

Another way we can perform cross-validation is to remove data at a sensor, and evaulate the predictions made. This will allow us to evaluate the imputation model's performance within sample, which is the primary concern with them imputation process.  
We select a sensor which has a small proportion of missing values, Bourke Street Mall (South), which has `r round((1 - na_prop["Bourke Street Mall (South)"])*100, 3)`% of data available. As observed in the data, missing data is present in different ways: random, short periods of missing data (no more that 6 hours) or long periods of missing data (multiple months). We will simulate both types of missingness at this sensor location.

@@

```{r, fig.width = 8, fig.height=2, fig.cap = "Plot of simulated missingness method 1 at Bourke Street Mall (South)"}
ggplot() + geom_line(aes(x = dfa2$Date_Time, y = mall_missing_1),
                     size = 0.1, alpha = 1) +
           labs(y = "Pedestrian Count", x = "Time and Date",
                title = "Simulated Missingness 1") +
          theme_minimal() + theme(plot.title = element_text(hjust = 0.5))
```

@@

```{r, fig.width = 8, fig.height=2, fig.cap = "Plot of simulated missingness method 2 at Bourke Street Mall (South) with 20% data missing"}
ggplot() + geom_line(aes(x = dfa2$Date_Time, y = mall_missing_2),
                size = 0.1, alpha = 1) +
           labs(y = "Pedestrian Count", x = "Time and Date",
                title = "Simulated Missingness 2") +
          theme_minimal() + theme(plot.title = element_text(hjust = 0.5))
```

The MARE of the estimated model is `r round(mall_mare_2*100, 2)`%, which is similar to what was obtained when the missing values were random. This demonstrates that the model quite well when we remove 20% of the observations from the training data.

```{r}
miss_seed <- min(max(sample(1:nrow(dfa2),1) - 6573, 0), 13146)
miss_idx <- seq(miss_seed, miss_seed + 13146)

mall_missing_3 <- dfa2$`Bourke Street Mall (South)`
mall_missing_3[miss_idx] <- NA

  sensor_df <- mall_missing_3 %>% unlist

  options(na.action = na.omit)
  mall_model_3 <- glm(sensor_df ~ Time*close_df_scale + Time*close_df_scale2,
               data = tr_dat, family = quasipoisson())
  options(na.action = na.pass)
  mall_fit_3 <- predict.glm(mall_model_3, newdata = tr_dat,
                            type = "response")
  options(na.action = na.omit)

  mall_mare_3 <- mean(abs(mall_fit_3 - unlist(dfa2$`Bourke Street Mall (South)`)),
                          na.rm = T) / mean(unlist(dfa2$`Bourke Street Mall (South)`),
                                            na.rm = T)

```{r, fig.width = 8, fig.height=2, fig.cap = "Plot of simulated missingness method 2 at Bourke Street Mall (South) with 50% data missing"}
ggplot() + geom_line(aes(x = dfa2$Date_Time, y = mall_missing_3),
                size = 0.1, alpha = 1) +
           labs(y = "Pedestrian Count", x = "Time and Date",
                title = "Simulated Missingness 3") +
          theme_minimal() + theme(plot.title = element_text(hjust = 0.5))
```

Removing 50% of the data using the same method, we find that the in-sample MARE is not affected very much by the significantly large proportion of missing training data where it is only `r round(mall_mare_3*100, 2)`%.

# Predictive Model

For the purpose of prediction, we want to use the most simplistic model possible to make quick predictions of the expected pedestrian counts at each location. The model should not be expected to be able to predict large deviations from time and date based estimates.

Similar to the simple imputation model used, we will also use a generalised linear model with quasi-poisson errors using time and date based variables as predictors. The model specification used is:

$$E \left[\text{HourlyCounts} \vert  \text{Sensor}, \text{Month}, \text{DayType}, \text{Time} \right] = \exp (\mu_{\text{Sensor}, \text{Month}, \text{DayType}, \text{Time}}) $$

When we estimate the GLM at each sensor location with R, we only need to save the elements in the model object which are required in order to retain the functionality of R's `predict.glm()` function. Saving the entire GLM object as given by the base R `glm()` function is highly inefficient when working with large datasets. This is due to the model object providing redundancies. In particular, elements which are unnecessarily saved for our purposes: the training data, the predictors in a model matrix format (the matrix of $\mathbf{X}$ after being formatted to be used for regression, such as recoding categorical variables into dummy variables), the response variable as another vector ($\mathbf{Y}$), the fitted values of the model after estimation, the residuals of the estimates, as well as the estimated effects.

The result of removing all the unnecessary redundant data elements in the GLM object is a storage size reduction from `r mall_model_2 %>% object.size() %>% format(units = "auto")` to `r mall_model_2 %>% stripGlmLR() %>% object.size() %>% format(units = "auto")`. This is only `r round((mall_model_2 %>% stripGlmLR() %>% object.size() %>% as.numeric/ mall_model_2 %>% object.size() %>% as.numeric)*100, 2)`% of the size of the original GLM object. With `r length(unique(ped_data$Sensor))` sensors (and models), this represents a reduction in model object size from `r format((mall_model_2 %>% object.size()) * length(unique(ped_data$Sensor)), units = "auto")` to `r format((mall_model_2 %>% stripGlmLR() %>% object.size()) * length(unique(ped_data$Sensor)), units = "auto")`.
